---
title: "Homework 1"
author: "Ethen Liu, Sai Haran, Sophia Hoffman, Annie Didier, Arindam Bhattacharya"
date: "1/10/2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Homework 1

```{r, message=FALSE, warning=FALSE}
library(lubridate)
library(data.table)

# change the working directory to where the dataset lies
setwd('/Users/ethen/Desktop/northwestern/winter/MSIA 421 Data Mining/hw1')
items <- fread('items.csv')
orders <- fread('orders.csv')

```

## Question 1 

How many unique customers.

```{r}

length( unique(orders$id) )

```

## Question 2 - 4

- 2. Suppose that the current date is 01JUL2013, compute the recency (most recent purchase)
- 3. Find the monetary value (total amount spent, item price times quantity) per customer
- 4. Find frequency (number of orders per customer)

Report the mean and standard deviation of recency, frequency, monetary value.

```{r, results='hide'}

orders[ , orddate := dmy(orddate) ]
rfm <- orders[ , .( recency = max(orddate),
					frequency = length(unique(ordnum)),
					monetary = sum(price * qty) ), by = id ]
rfm[ , recency := as.numeric( dmy('01JUL2013') - recency ) ]

compute_mean_and_sd <- function(value) {
	# pass in value to compute the mean and standard deviation,
	# the outputed list will contain the passed in variable
	# as an indication of what value is being computed
	value_name <- deparse( substitute(value) )
	mean_name <- paste('mean', value_name, sep = '_')
	sd_name <- paste('sd', value_name, sep = '_')
	info <- list( mean(value), sd(value) )
	names(info) <- c(mean_name, sd_name)
	return(info)
}

```

```{r}

recency  <- rfm[['recency']]
compute_mean_and_sd(recency)

monetary <- rfm[['monetary']]
compute_mean_and_sd(monetary)

freqency <- rfm[['frequency']]
compute_mean_and_sd(freqency)

```

## Question 5

Compute the number of unique items that each customer purchased in each category and report the mean and standard deviation for each category.

```{r}
merged <- merge(items, orders, by = 'sku')
category_counts <- merged[ , .( counts = length( unique(name) ) ), by = .(id, category) ]
counts <- category_counts[['counts']]
category_counts[ , compute_mean_and_sd(counts), by = category ]
```

## Question 6

Compute the entropy and examine if those with higher entropy have more diversity in their reading interest.

```{r}
# compute the entropy and sort them in decreasing order
entropy <- category_counts[ , {
	p <- counts / sum(counts)
	entropy <- -sum( p * log10(p) )
	list(entropy = entropy)
}, by = id ][order(-entropy), ]
```


```{r}
# the top and bottom entropy's customer information
category_counts[ id == 4335961, ]
category_counts[ id == 4313828, ]
```

Base on the printed result, the customer that has the highest entropy does in fact have a more diverse taste than the customer that has the lowest entropy.

