---
title: "Data Mining HW 3"
author: "Ethen Liu, Sai Haran, Sophia Hoffman, Annie Didier, Arindam Bhattacharya"
date: "January 23, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1

a)
```{r}
X = matrix(c(1,3,5,0,1, 2,4,4,2,3, 3,5,3,4,5), nrow=5)
x_tran_x = t(X) %*% X
x_x_tran = X %*% t(X)
x_tran_x
x_x_tran
```

b)
```{r}
ei_x_tran_x = eigen(x_tran_x)
ei_x_x_tran = eigen(x_x_tran)
ei_x_tran_x
ei_x_x_tran
```

c)
```{r}
A = ei_x_tran_x$vectors%*%diag(ei_x_tran_x$values)%*% t(ei_x_tran_x$vectors)
A
```
The output is the same as 2a. 

d)
```{r}
svd_x = svd(X)
#Compare to b
svd_x$v
ei_x_tran_x$vectors
svd_x$u
ei_x_x_tran$vectors
(svd_x$d)^2
(ei_x_tran_x$values)
ei_x_x_tran$values
```
We can see that V gives the eigenvectors of X'X and that the first two columns of U correspond to the first two eigenvectors of XX' since the rank of X is 2.  

e)
```{r}
ULV = svd_x$u %*% diag(svd_x$d) %*% t(svd_x$v)
ULV
X
```

f)
```{r}
#set the third value to 0 since X has rank 2
svd_x_rank2 = svd_x$d[1:2] 
#set the smallest eigenvalue to 0 to do the 1-d projection
svd_x_oned = svd_x$d
svd_x_oned[2:3] = 0
svd_u_one = svd_x$u
svd_u_one[,2:3] = 0
svd_v_one = svd_x$v
svd_v_one[,2:3] = 0
xhat = svd_u_one %*% diag(svd_x_oned) %*% t(svd_v_one)
xhat
```
xhat is very close to x, so the 1D estimate is a good approximation. 

g)
```{r}
ssxhat = sum(xhat^2)
ssxhat
svd_x$d[1]^2
```
The Frobenius norm, or one dimensional variance, of xhat is the same as the square of the first singular value. 

h)
```{r}
SSE = sum((X-xhat)^2)
SSE
svd_x$d[2] ^ 2
```

It is equal to the square of the k + 1 singular values, where k is the number of the top k singular value we chose to compute the Xhat approximation.

i) 
```{r}
energy = svd_x_oned^2/(sum(svd_x$d^2))
energy
```

# Question 2

```{r, message=FALSE, warning=FALSE, results='hide'}

library(data.table)
setwd("/Users/ethen/Desktop/northwestern/winter/MSIA 421 Data Mining/hw3")
theater <- fread('theater.csv')
theater <- theater[ , -c(3, 5, 7, 10), with = FALSE ]

# the theater subset data will be used in question e
theater_subset <- theater[ , .(dinner, play) ]
theater[ , c('dinner', 'play', 'age', 'educ', 'income', 'cnty') := NULL ]

```

a)
```{r, message=FALSE, warning=FALSE}
alpha_theater1 <- psych::alpha(theater, check.keys = TRUE)
alpha_theater1
```

b)
```{r}
pca1 <- prcomp(theater)
pca1$sdev
```

4 eigenvalues are greater than 1.

```{r}
pca1$rotation[, 1]
```

noteduc, dislike have smaller first loading vectors when it comes to magnitude.

c)

```{r, results='hide'}

# after removing noteduc and dislike, 
# the drop of reliability remains the same for
# every single variable that we dropped
theater[ , c('dislike', 'noteduc') := NULL ]
alpha_theater2 <- psych::alpha(theater, check.keys = TRUE)
alpha_theater2

```

```{r}
# and the explained ratio of the first loading vector is larger
pca2 <- prcomp(theater)
summary(pca1)
summary(pca2)
```


d)

```{r, results='hide'}
# reverse code the reversed coded columns
reverse_col <- c('irritate', 'bad', 'cannotappreciate')
max_score <- 7
theater[ , (reverse_col) := lapply(.SD, function(col) {
	(max_score + 1) - col
}), .SDcols = reverse_col ]

# compute attitude variable
theater_subset[ , attitude := rowSums(theater) / ncol(theater) ]
```

```{r}
theater_subset
```


e)
```{r, results='hide'}
# there are NA values in theater subset, we tried dropping the
# NA values and setting the NA values to 0 and obtained similar result
theater_subset1 <- theater_subset[ complete.cases(theater_subset), ]
theater_subset1[ , new_feature := log(dinner + play + 1) ]

theater_subset[ is.na(dinner), dinner := 0 ]
theater_subset[ is.na(play), play := 0 ]
theater_subset[ , new_feature := log(dinner + play + 1) ]
```

```{r}
with( theater_subset1, cor(new_feature, attitude, method = 'pearson') )
with( theater_subset, cor(new_feature, attitude, method = 'pearson') )
```


f)

The sum of dinner and play is not within the normal range of 1 to 7, by taking the log, we transform it to a narrower and more comparable range.

# Question 3

```{r}
music <- read.csv("music.csv")
retain_cols <- c('V28', 'V29', 'V30', 'V31', 'V32', 'V33', 
                 'V34', 'V43', 'V46', 'V47', 'V50', 'V52')
music <- music[, retain_cols]
```

a)
```{r}
alpha_music <- psych::alpha(music)
alpha_music
```

b)
```{r}
pca <- prcomp(music)
pca$sdev
```

3 eigenvalues are greater than 1.

c)

A person's similarity of taste in music with friends or a person's willingless to share his/her personal taste in music.


